---
title: "specimens_workspace"
output:
  html_document: default
  pdf_document: default
  word_document: default
date: "2024-01-25"
---
R markdown automatic settings
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Working with specimen database:
Note that the database is limited (lack of responses, for example Rotterdam only gave us species names but not numbers) and extremely biased
```{r}
#setting the basics and importing file
library(tidyverse)
library(readr)

#This below part is personal
library(readr)
toothedwhale_specimens <- read_delim("C:/Users/Beau Luiten/Downloads/toothedwhale_specimens.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)
View(toothedwhale_specimens)
```

Filtering out irrelevant information for R analysis and preparing database for analysis
```{r}
#Subset of only relevant specimen information
#leaving out things like notes, contact persons
specimens <- select(toothedwhale_specimens, source_code, genus,
                    species, common_name, type, date, count, town, 
                    country, notes) 
specimens$date <- as.numeric(specimens$date)
group_by(specimens, species)
```

```{r}
#Filtering for dated specimens
specimens_dated <- filter(specimens, !is.na(date))
View(specimens_dated)
```

In case someone in the future wants to work with Physeter and phocoena only
```{r}
#Grouping by species and selecting only sperm whale and harbor porpoise
specimens_phocoena_physeter <- filter(specimens, 
                                      genus==c("Physeter",
                                               "Phocoena"))
specimens_phocoena_physeter
```

Next I want to plot a species distribution. 
By making it into a function, it can be applied to various databases: e.g. to compare the distribution between dated and non-dated specimens
```{r Species distribution graphs}
library(ggplot2)
# Function to generate species distribution plot
plot_species_distribution <- function(data, title) {
  plot <- data %>% 
    ggplot(aes(x = common_name)) +
    geom_bar(stat = "count", colour = "black", fill = "white") +
    geom_text(stat = "count", aes(label = ..count..), vjust =
                -0.5, size = 4) +  # Add count labels
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
    labs(title = title, y = "Count", x = "Species")
  return(plot)
}

#Species distribution for all specimens
species_distribution <- plot_species_distribution(specimens, 
                        "Species distribution") +
  ylim(0,300) #subject to change in future, also for next plots
print(species_distribution)

#Species distribution for dated specimens
species_distribution_dated <- plot_species_distribution(specimens_dated,
              "Species distribution of dated specimens") +
  ylim(0,50)
print(species_distribution_dated)
```

When combined into a single figure both can directly be compared
However, it is more difficult to interpret especially for the less frequent species
```{r Species distribution combined graph}
install.packages("dplyr")
library(dplyr)

# Function to generate species distribution plot
plot_species_distribution <- function(data, data_dated, title) {
  # Combine datasets and create a new column to distinguish
  combined_data <- bind_rows(
    mutate(data, dataset = "All Specimens"),
    mutate(data_dated, dataset = "Dated Specimens"))
  
  # Plot
  plot <- combined_data %>% 
    ggplot(aes(x = common_name, fill = dataset)) +
    geom_bar(position = "dodge", colour = "black") +
    geom_text(stat = "count", aes(label = ..count..), position = position_dodge(width = 0.9), vjust =
                -0.5, size = 3.5) +  # Add count labels
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
    labs(title = title, x = "Species", y = "Number of specimens") +
  scale_fill_manual(values = c("All Specimens" = "#E1CC75", "Dated Specimens" = "#cd393e"))
  return(plot)
}

#Species distribution for combined data
species_distribution_combined <- plot_species_distribution(specimens,
                        specimens_dated,                                                  "Species distribution") +
  ylim(0,300)
print(species_distribution_combined)
```

Another way is to show the information in a table
```{r Species distribution table}
library(dplyr)

species_counts <- specimens %>%
  group_by(common_name) %>%
  summarise(total_count = n(),
            dated_count = sum(!is.na(date))) %>%
  mutate(percentage_dated = (dated_count / total_count) * 100, digits = 1) %>%
  rename(Species = common_name,
         `Number of specimens` = total_count,
         `Number of dated specimens` = dated_count,
         `Percentage dated` = percentage_dated) %>%
arrange(desc(`Number of specimens`))
print(species_counts)

```

```{r}
#Make sure to have these packages installed
library(dplyr)
library(knitr)
library(kableExtra)

#Code to produce a table for species distribution
species_counts <- specimens %>%
  group_by(common_name) %>%
  summarise(total_count = n(),
            dated_count = sum(!is.na(date))) %>%
  mutate(percentage_dated = (dated_count / total_count) * 100) %>%
  rename(Species = common_name,
         `Number of specimens` = total_count,
         `Number of dated specimens` = dated_count,
         `%` = percentage_dated) %>%
  arrange(desc(`Number of specimens`))

#Using kableExtra to make it a table suitable for publication
kable(species_counts, format = "html", digits = 1, caption = "Species Distribution") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

```{r Age distribution graph}
#Plotting age distribution of dated specimens
age_distribution <- specimens_dated %>% 
  ggplot(aes(x=`date`)) +
  geom_histogram(stat="count", colour="#7d9b76", fill="#7d9b76",
                 binwidth = 1) +
  xlim(1600, 2020) + #as our dated ones range in this period now only, this shows the distribution on this small scale best
  labs(title = "Age distribution of dated specimens", x = "Date", y = "Number of specimens", size = 15)

# Add a density line (could be left out)
age_distribution <- age_distribution +
  geom_density(aes(y = ..count..), color = "#cd393e")
age_distribution

```

Distribution of specimens per country, and fraction that is dated
```{r Location distribution graph}
country_counts <- specimens %>%
  group_by(country) %>%
  summarise(total_count = n(),
            dated_count = sum(!is.na(date))) %>%
  mutate(fraction_dated = dated_count / total_count)

# Create bar plot
ggplot(country_counts, aes(x = country)) +
  geom_bar(aes(y = total_count), stat = "identity", color = "#7d9b76", fill = "white") + 
  geom_bar(aes(y = fraction_dated * total_count), stat = "identity", color = "#cd393e", fill = "#cd393e", alpha = 0.7) + 
  geom_text(aes(label = total_count, y = total_count), vjust = -0.5, size = 6) + #only total number
  labs(title = "Distribution of specimens per country", y = "Count of specimens", x = "Country") + 
  ylim(0,300) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=14))
```
